Fourier-domain modulations and delays of gravitational-wave signals

### 2017-03-28 docs/paper_method.tex
Edits to the paper.
Also added more TDItags allowed for SNR computation.

\title{Modulations of gravitational-wave signals: a Fourier-domain approach for the response of LISA-type detectors and for precessing binaries}

@@

We present a Fourier-domain based approach to the problem of the modulation of gravitational wave signals. The two main contexts where modulations are important for gravitational-wave astronomy: (i) the time-dependent response of space-based instruments such as LISA; and (ii) the transposition in Fourier domain of the frame decomposition of precessing waveforms, where the precessing waveform is approximated by rotating a non-precessing waveform to follow the precession of the orbital plane. Our results recover and generalize previous approaches based on an extension of the stationary phase approximation, and can be applied to full inspiral-merger-ringdown signals. In the case of the response LISA-type detectors, we show that higher-order corrections remain small except at the two extremes of the detector's frequency band, and we show that including them reduce the errors. In the case of precessing binaries, we perform a limited exploration of the merger-ringdown range based on a toy model for the precession. We find that, depending on the post-merger frame velocity, higher-order corrections can become quantitatively large and even challenge our perturbative formalism. We develop a direct convolution approach to capture these post-merger features. However, in that case these corrections appear mainly as amplitude corrections and give raise to a small unfaithfulness.

@@

With the recent detections of gravitational-wave signals emitted by the coalescences of binary black holes, announced by the LSC collaboration after the first obervation run of the two LIGO instruments~\cite{LIGO-theevent-2016,LIGO-christmasevent-2016,LIGO-O1BBH-2016}, gravitational-wave astronomy has entered its observational era. With the recent start of LIGO's second observation run, and later the extension of the ground-based detectors network with Virgo~\cite{Virgo}, KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}, many more detections from compact object coalescences are to come, with binary black holes being expected to be the dominant sources~\cite{}. Moreover, the European Space Agency has issued a mission call for its L3 mission, adressing the science theme ``The Gravitational Universe'' with a tentative launch around 2034, possibly opening the way for a space-borne observatory based on the concept of the Laser Interferometer Space Antenna (LISA)~\cite{LISA09}. Among other sources of gravitational waves, LISA will be able to detect and characterize binary black hole coalescences to cosmological distances over a wide range of masses, from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.

Data analysis for gravitational-wave observations of compact binary coalescences require accurate models (or templates) for the signals, both for ensuring efficient detections of signals that may be buried in instrumental noise, and to extract the physical parameters of the source in subsequent analysis. This has prompted an important and ongoing effort to improve gravitational wave templates. In parallel of the improvement of the accuracy of the templates, an crucial element of the analysis is the computational efficiency of template generation, as Bayesian analysis for parameter estimation of gravitational-wave signals may require millions of evaluations of the likelihood function to sample the posterior probability distribution. Two important axis of improvement of the templates are the coverage of all phases of the evolution of the system, with the inspiral, merger and ringdown (IMR), and the inclusion of the effects of the spins of the compact objects.

State-of-the-art IMR templates need to combine information from the perturbative results of post-Newtonian theory covering the inspiral (see e.g.~\cite{BlanchetLiving}) and from numerical relativity simulations covering the end of the inspiral and the merger-ringdown phase (see e.g.~\cite{Pfeiffer12}). The Effective-One-Body (EOB) approach~\cite{BD99} does so by incorporating a resummation of post-Newtonian results and a calibration to numerical relativity, and has been extended to include both spins of the two bodies~\cite{Taracchini+13, Pan+13, Bohe+16}. Phenomenological templates, which extend post-Newtonian signals by postulating a Fourier-domain analytic expression for the amplitude and phase and by fitting coefficients against numerical relativity waveforms~\cite{Husa+15, Khan+15}, and can include a simplified model for precession~\cite{Hannam+13}. The second type of models produce directly the amplitude and the phase of the Fourier-domain waveform in the form of a piecewise analytic fit. EOB waveforms, by contrast, require to integrate ordinary differential equations in the time-domain, which can be time-consuming. For the non-precessing case, Reduced Order Models (ROM), also called surrogate models, have been developed to speed up considerably the waveform generation, without losing accuracy~\cite{Field+13, Puerrer14, Bohe+16}. Both approaches are therefore able to produce efficiently IMR waveforms. An important aspect is that the resulting waveforms can be represented in the Fourier domain by an amplitude and phase for each mode, with only a few hundred samples~\cite{Puerrer14}.

A complete representation of the spin effects in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure a precession of the orbital plane as it evolves, leading to modulations of the signal~\cite{ACST94}. Precessing EOB templates can represent all six degrees of freedom of the two spins~\cite{Pan+13, BTB16} but still require a costly time-domain integration; while precessing phenomenological templates~\cite{Hannam+13} decrease the dimensionnality by using effective spin parameters representing only the dominant effects of the spins. To model the effect of precession on the emitted waveform, it has been proposed~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} to decompose precessing waveforms by performing a time-dependent rotation, following the precession of the orbital plane. The resulting waveform in the rotated frame is then close to a non-precessing waveform, an approximation that is used in the construction of the inspiral part of precessing EOB waveforms~\cite{Pan+13} and in the construction of precessing phenomenological waveforms~\cite{Hannam+13}. For Fourier-domain waveforms, one needs to understand how to translate the time-domain modulation created by the frame rotation into a Fourier-domain modulation.

The Stationary Phase Approximation (SPA) (see e.g.~\cite{Thorne300, CF94}) is a common and very good approximation to compute the Fourier transform of non-precessing signals in the inspiralling phase, but applying it directly to the modulated signal is prone to pathologies. In Ref.~\cite{KCY13,KCY14}, a formalism (called shifted uniform asymptotics or SUA) was introduced to go beyond the SPA and compute more accurately the modulation in the Fourier domain; however, this formalism still relies on the SPA for the underlying signal, and is as such limited to inspiralling signals. Another intuitive approach consists in using a time-to-frequency correspondence, such as the one given by the SPA, to directly carry over the modulation to the Fourier domain. This corresponds, in a sense to be detailed below, to the zeroth-order approximation of the SUA, and it is used implicitly\footnote{With the difference that directly frequency-dependent expressions for the modulation are used, which are then extended to also cover high frequencies in the merger part of the signal, as we will discuss below.} for the simplified precession of the phenomenological waveforms of Ref.~\cite{Hannam+13}. Consistently extending these previous approaches to the merger-ringdown part of the signal is one of the objectives of this paper.

Prospective parameter recovery for LISA observations of comparable-mass binary black hole coalescences is another frontier of gravitational wave signal modelling, becoming a very timely topic with the upcoming effort for the ESA L3 mission characterization in the next few years. In the past, many LISA data analysis studies used mainly post-Newtonian, inspiral-only models, and were limited to the Fisher-matrix formalism instead of full Bayesian analysis (among numerous works, see e.g. \cite{Cutler97,LISAPE09}). However, the Fisher matrix analysis, although considerably cheaper to perform, has known shortcomings including at high SNR~\cite{Vallisneri08,Cokelaer08,Rodriguez+13}. In Ref.~\cite{Klein+15}, an improved model was used, that included the effects of both spins including the corrections beyond the SPA investigated in~\cite{KCY14}; the limitation to inspiral-only waveforms was compensated by SNR reweighting using full IMR waveforms.

The response of the LISA instrument, by contrast with the response of ground-based detectors, has the specificity of being time-dependent. For a signal accumulating over several months, the motion and change of orientation of the detector constellation along its orbit lead to time variability in the response of the detector, which is crucial for the localization of the gravitational-wave source in the sky. More precisely, these effects can be separated as a modulation coming from the rotation of the constellation, and a delay coming from the motion of the spacecrafts in the wavefront of the incoming wave.

The purpose of this paper is to propose a formalism allowing to transpose directly in the Fourier domain the processing of a signal through a time-domain modulation and delay, while retaining the compactness of a Fourier-domain amplitude and phase representation of the signals. This allows us to address both the issue of a direct Fourier-domain response of the LISA instrument, as well as the issue of Fourier-domain precession modulation for IMR signals.

[plan]
The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, we will introduce our notations and conventions for Fourier-domain waveforms, motivate the problem of Fourier-domain modulation and give the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, and introduce new timescales based on the Fourier-domain signal. In Sec.~\ref{sec:quadphasedelay}, we refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays, and make the connection to previous approaches of the problem. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.

>>>>

\title{Fourier-domain modulations and delays of gravitational-wave signals}

@@

We present a Fourier-domain based approach modulations and delays of gravitational wave signals, a problem which arises in two different contexts. For space-based detectors like LISA, the orbital motion of the detector introduces a time-dependency in the response of the detector, consisting of both a modulation and a varying delay. In the context of signals from precessing spinning binary systems, a useful tool for building models of the waveform consists in representing the signal as a time-dependent rotation of a quasi-non-precessing waveform. In both cases, being able to compute transfer functions for these effects directly in the Fourier domain would allow to gain performance for data analysis applications by using fast frequency-domain waveforms. Our results recover previous approaches based on an extension of the stationary phase approximation, generalize them by including delays and higher-order corrections, and can in principle be applied to full inspiral-merger-ringdown signals. In the LISA case, we find that higher-order corrections remain small except at the two extremes of the detector's frequency band, where we show that including them reduces the errors. In the case of precessing binaries, we perform a limited exploration of the merger-ringdown range based on a toy model for the precession. We find that, depending on the post-merger frame velocity, higher-order corrections can become quantitatively large to the point of challenging our perturbative formalism, while affecting mainly the amplitude and giving raise to a limited unfaithfulness. We develop an alternative direct convolution approach to represent accurately these post-merger features.

@@

With the recent detections of gravitational-wave signals emitted by the coalescences of binary black holes, announced by the LVC collaboration after the first obervation run of the two LIGO instruments~\cite{LIGO-theevent-2016,LIGO-christmasevent-2016,LIGO-O1BBH-2016}, gravitational-wave astronomy has entered its observational era. With the recent start of LIGO's second observation run, and later the extension of the ground-based detectors network with Virgo~\cite{Virgo}, KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}, many more detections from compact object coalescences are to come, with binary black holes being expected to be the dominant sources~\cite{}. Moreover, the European Space Agency has issued a mission call for its L3 space mission with a tentative launch around 2034, adressing the science theme ``The Gravitational Universe'', possibly opening the way for a space-borne observatory based on the concept of the Laser Interferometer Space Antenna (LISA)~\cite{LISA09,LISA17}. Among other sources of gravitational waves, LISA will be able to detect and characterize comparable-mass binary black hole coalescences to cosmological distances over a wide range of masses, from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.

Data analysis for gravitational-wave observations of compact binary coalescences require accurate models (or templates) for the signals, both for ensuring efficient detections of signals that may be buried in instrumental noise, and to extract the physical parameters of the source in a subsequent analysis. This has prompted an important and ongoing effort to improve gravitational wave templates. Two important qualitative features for the improvement of the templates are the coverage of all phases of the evolution of the system, including the inspiral, merger and ringdown (IMR), which became possible thanks to the breakthrough of numerical relativity~\cite{Pretorius05,Baker+06,Campanelli+06}, and the inclusion of the effects of the spins of the compact objects, that can lead to a precession of the orbital plane~\cite{ACST94,K95}. Besides accuracy, the computational efficiency of template generation is crucial, as Bayesian analysis for parameter estimation of gravitational-wave signals, as was performed for the LIGO detections~\cite{LIGO-theeventPE-2016,LIGO-O1BBH-2016}, may require millions of evaluations of the likelihood function to sample the posterior probability distribution.

State-of-the-art IMR templates need to combine information from the perturbative results of post-Newtonian theory covering the inspiral (see e.g.~\cite{BlanchetLiving}) and from numerical relativity simulations covering the end of the inspiral and the merger-ringdown phase (see e.g.~\cite{Pfeiffer12}). The Effective-One-Body (EOB) approach~\cite{BD99} does so by incorporating a resummation of post-Newtonian results and a calibration to numerical relativity, and has been extended to include both spins of the two bodies~\cite{Taracchini+13, Pan+13, Bohe+16}. Phenomenological templates extend post-Newtonian models by postulating an analytic ansatz for the Fourier-domain amplitude and phase, fitting coefficients against numerical relativity waveforms~\cite{Husa+15, Khan+15}, and can include a simplified model for precession~\cite{Hannam+13}. The second type of models produce directly the amplitude and the phase of the Fourier-domain waveform in the form of a piecewise analytic function. EOB waveforms, by contrast, require to integrate ordinary differential equations in the time-domain, which can be time-consuming. For the non-precessing case, Reduced Order Models (ROM), also called surrogate models, have been developed to speed up considerably the waveform generation, without losing accuracy~\cite{Field+13, Puerrer14, Bohe+16}. Both approaches are therefore able to produce efficiently IMR Fourier-domain waveforms. An important aspect is that the resulting waveforms can be represented by an amplitude and phase for each mode, with only a few hundred samples~\cite{Puerrer14}.

A complete representation of the spin effects in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure a precession of the orbital plane as it evolves, leading to modulations of the signal as seen by the observer~\cite{ACST94}. Precessing EOB templates incorporate all six parameters of the two spins~\cite{Pan+13, BTB16} but still require a costly time-domain integration; while precessing phenomenological templates~\cite{Hannam+13} decrease the dimensionnality by using effective spin parameters representing only the dominant effects of the spins. To model the effect of precession on the emitted waveform, it has been proposed~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} to decompose precessing waveforms by performing a time-dependent rotation, following the precession of the orbital plane. The resulting waveform in the rotated frame can then be modelled by a non-precessing waveform, an approximation which is used both in the construction of the inspiral part of precessing EOB waveforms~\cite{Pan+13} and in the construction of precessing phenomenological waveforms~\cite{Hannam+13}. To follow this modelling approach and create efficiently Fourier-domain waveforms, one needs to understand how to translate the time-domain modulation created by the frame rotation into a Fourier-domain transfer function, which is our first motivation.

Prospective parameter recovery studies for LISA observations of comparable-mass binary black hole coalescences aim at characterizing the scientific performance of the future experiment. They become a timely topic with the upcoming effort for the ESA L3 mission characterization in the next few years. However, the recent progresses in waveform modelling, most notably the availability of cheap IMR waveforms, have not yet made their way to the field of LISA data analysis studies. In the past, such studies used mainly post-Newtonian, inspiral-only models, and were limited to the Fisher-matrix formalism instead of full Bayesian analysis (among numerous works, see e.g. \cite{Cutler97,LISAPE09}). Refs.~\cite{McWilliams+09,McWilliams+10,McWilliams+11} explored the effects of including the merger and ringdown, keeping the mass parameters fixed. However, the Fisher matrix analysis, although considerably cheaper to perform, is valid only in the limit of a high signal to noise ratio (SNR) and has known shortcomings~\cite{Vallisneri08}. Several implementations of Bayesian analyses were developed~\cite{CP06,Wickham+06,Babak+08,GP09,Feroz+09,Petiteau+10,PC13} (see also~\cite{MLDC4} and references therein), with Ref.~\cite{PC15} proposing a direct comparison of Fisher and Bayes analyses, but they were limited to inspiral-only waveforms. One exception was Ref.~\cite{Babak+08} where IMR signals were used in a Bayesian study, but there the masses of the black holes were kept fixed. In the recent Ref.~\cite{Klein+15}, the limitation to inspiral-only precessing post-Newtonian waveforms was compensated by an SNR reweighting procedure. Thus, due mainly to their high computational cost, full-fledged Bayesian parameter estimation simulations for IMR signals are still missing.

By contrast with ground-based detectors, the response of a LISA-type instrument has the specificity of being time-dependent~\cite{Cutler97}. For signals possibly accumulating over several months, the motion and change of orientation of the detector constellation along its orbit lead to significant time variability in the form of a modulation and a varying delay, which contain information about the localization of the gravitational-wave source in the sky. This complicated detector response increases the computational cost of parameter-estimation analyses. To leverage the performance of state-of-the-art Fourier-domain IMR templates~\cite{BTB16,Khan+15}, one needs to be able to process the signals through the time-dependent response of the dectector while staying in the Fourier domain, which is our second motivation.

The purpose of this paper is to propose a formalism allowing to transpose directly in the Fourier domain the processing of a signal through a time-domain modulation and delay, while retaining the compactness of a Fourier-domain amplitude and phase representation of the signals. This will allow us to address both the issue of the Fourier-domain response of the LISA instrument, as well as the issue of Fourier-domain precession modulation for IMR signals from precessing binaries.

In previous works, the Stationary Phase Approximation (SPA) (see e.g.~\cite{Thorne300, CF94}) has been often used for this purpose. The SPA is a common approximation to compute the Fourier transform of non-precessing signals during the inspiralling phase. In the case of precessing binaries, applying it directly to the modulated signal is however prone to pathologies. In Ref.~\cite{KCY13,KCY14}, a formalism (called shifted uniform asymptotics or SUA) was introduced to go beyond the SPA and compute more accurately the modulation in the Fourier domain; however, this formalism still relies on the SPA for the underlying precessing-frame signal, and is as such limited to inspiralling signals. Another intuitive approach consists in using a time-to-frequency correspondence, such as the one given by the SPA, to directly carry over the modulation to the Fourier domain. This corresponds, in a sense to be detailed below, to the zeroth-order approximation of the SUA, and it is used implicitly\footnote{With the difference that directly frequency-dependent expressions for the modulation are used, which are then extended to also cover high frequencies in the merger-ringdown part of the signal, as we will discuss below.} for the simplified precession of the phenomenological waveforms of Ref.~\cite{Hannam+13}. In the case of the LISA response, the SPA provides a natural map from time-domain (for the orbit) to Fourier-domain (for the signal)~\cite{Cutler97}. Ref.~\cite{Klein+15} included the orbital motion of the detector in the SUA treatment. Consistently extending these previous approaches to the merger-ringdown part of the signals, including the delays in the LISA case, is part of the objectives of this paper.

The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, we will introduce our notations and conventions for Fourier-domain waveforms, motivate the problem of Fourier-domain modulation and introduce the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, and introduce new timescales based on the Fourier-domain signal. In Sec.~\ref{sec:quadphasedelay}, we refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays, and make the connection to previous approaches of the problem. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.
